---
output:
  html_document: default
  pdf_document: default
---
# **Analysis of methylation levels in acute lymphatic leukemia versus B-cell lines**

## by Wiebke Albers, David Steeb, Nina Winkler, Lilian Dorner

### Introduction

**DNA methylation**

One of the cells regulatory mechanisms for switching genes on and off is methylation of cytosines in promotor or coding regions. The enzymes responsible for this methylation only recognize cytosines which are followed by a guanosine nucleotide in the strand. This conformation is called a CpG (cytosine phosphate guanosine). Upon methylation of the CpGs in promotor regions the transcription machinery is blocked. It can therefore not bind to its activating site which leads to downregulation or silencing of the downstream gene or genes. Methylation within a gene can be interpreted similarly. If an abundant amount of methylation is present within a gene sequence the RNA polymerase struggles to transcribe the DNA and is more likely to stop along the gene. This also leads to a downregulation of the said gene.

These CpGs are normally methylated. The exception to this are CpG-rich promoters called CpG islands which are rarely methylated. These CpG islands are often found as promoters of housekeeping genes which rarely ever show any methylation. In cancer however an epigenetic change in these CpG islands for example can lead to a major driving force for cancerous development. Hypermethylation of tumoursupressorgenes or hypomethylation of oncogenes inevitably favour cancer growth.

**Tumour type**

The type of cancer we are focussing on is acute lymphoblastic leukemia (ALL) arising from the B-cell line, which is the most common form with 75% of ALL cases (Terwillinger and Abdul-Hay, 2017). It is a very aggressive form of leukemia. Approximately 60% of leukemia-diseases in children under 20 years are due to ALL making it the predominant form of leukemia (Pui *et al.*, 2011). Combining this knowledge with methylation status our project is comparing genome-wide methylation data of diseased and healthy patients. The samples were either taken from the bone marrow (healthy and diseased) or the tonsils (healthy) of 0-18 year old patients.

**Biological question - goal**

wip

### Step1 - Quality control

Before starting our analysis the dataset first has to undergo different stages of data cleaning and quality control. In order to do that we need to remove any non significant or ambiguous rows and remove or replace all NAs. As a last step our Beta-values are converted into M-values for a statistically more accurate representation of the methylation levels.

**Trimming the dataset and creation of matrices for later use**

The columns "Strand", "symbol", "GC", "C" und "G" are not relevant for our analysis and are therefore removed.
```{r Reading and trimming}
ALLBcell <- readRDS("ALL-Bcell_list.RDS.gz")
ALLpromotor <- ALLBcell[["promoters"]] 
ALLpromotor <- ALLpromotor[,c(-4,-5,-6,-8,-9,-10)]
```

The coverage columns were isolated in a seperate dataframe which was called ALLpromotorCov. Next the means of the coverage of the promoters were calculated and stored in ALLCovMeans. Finally these values were logarithmically converted to obtain a simpler plot.
```{r Creating matrices}
ALLpromotorCov <- ALLpromotor[,c(15:24)]
ALLCovMeans <- rowMeans(ALLpromotorCov)
ALLCovMeansLog <- log(ALLCovMeans)
```


**Plotting the histograms for means**

Plotting the means of the coverage values against their frequency.
```{r Plotting normal}
hist(ALLCovMeans, xlab = "Means of coverage values", ylab = , xlim = c(0,10000), ylim = c(0,4000), breaks = 1000, lwd = 1)
```

Plotting the logarithmic means of the coverage values against their frequency.
```{r Plotting logarithmic}
hist(ALLCovMeansLog, xlab = "Logarithmic means of coverage values", ylab = , xlim = c(0,13), ylim = c(0,250), breaks = 1000, lwd = 1)
```


**Finding a suited threshold**

The coverage equates to the average of the reads per base on any given gene region. During sequencing each base is determined by the majority of its reads. If there are very few (e.g. <25) reads for a single base or region it is prone to error due to statistical insecurity. The lower the amount of reads the higher are the chances that a few "wrong" reads alter the result of the read out base.
If there are very many (e.g. >10000) reads per base the result could be altered due to an amplification error during the PCR. This error would then be amplified as well if it appears early duing the amplification process. 
Therefore it is important to find a suited threshold for the coverages in order to assure the significance of the Beta-values which are directly depending on the sequencing accuracy.
Upon literature reseach we defined 14 promoters for genes (CDH1, CDKN2A, CDKN2B, CDKN1C, KLK10, DKK3, CDH13, PYCARD, DAPK1, PARKIN, PTEN, p73, APAF1, LATS1) as "relevant" for our tumour type (Roman-Gomez *et al.*, 2004). In order to find an upper threshold we took several factors into account. First we checked how many of the "relevant" promoters were removed for different thresholds. Even with a threshold of 99% of our coverage 3 promoters for genes (p73, APAF1, LATS1) always only had NA for the coverage values (any value outside of thresholds = NA). 
Next we checked how many "non-relevant" regions were removed while still keeping as many "relevant" promoters additionally reducing the coverage as much as needed to avoid PCR mediated errors.
Finally we agreed on a lower threshold of 1.5 % (28) and an upper threshold of 98.5 % (7753) of the coverage values since only 10 % of "non-relevant" data and 2 additional "relevant" promoter regions (p57, PTEN) were rejected. It formed the best compromise between the aforementioned factors.
```{r Checking for quantiles}
quantile(ALLCovMeans, probs = c(0.05, 0.95),na.rm = TRUE)
quantile(ALLCovMeans, probs = c(0.015, 0.985),na.rm = TRUE)
ALLpromotorCov1 <- ALLpromotorCov 
```

Next any values of the coverage above and below the thresholds were replaced with NA.
```{r Replacing outliers with NA}
ALLpromotorCov1[ALLpromotorCov > 7753] <- NA
ALLpromotorCov1[ALLpromotorCov1 < 28] <- NA
```


**Isolation of "relevant" promoters before and after threshold cleanup**

 The presence of the "relevant" promoters was checked and prepared for later calculation for the rejected amount of coverage values.
```{r Isolating "relevant" promoters}
ALLCovGen <- ALLpromotorCov[c("ENSG00000039068","ENSG00000078900","ENSG00000147889","ENSG00000147883","ENSG00000129757","ENSG00000129451","ENSG00000050165","ENSG00000140945","ENSG00000103490","ENSG00000120868","ENSG00000196730","ENSG00000185345","ENSG00000131023","ENSG00000171862"),]
ALLCov1Gen <- ALLpromotorCov1[c("ENSG00000039068","ENSG00000078900","ENSG00000147889","ENSG00000147883","ENSG00000129757","ENSG00000129451","ENSG00000050165","ENSG00000140945","ENSG00000103490","ENSG00000120868","ENSG00000196730","ENSG00000185345","ENSG00000131023","ENSG00000171862"),] 
```


**Checking for percentage of rejected coverage values**

Finally upon each tested threshold the amout of rejected values was calculated by subtracting the percentage of non NA values after threshold cleanup from 1. This calculation was always done for both "non-relevant" and "relevant" promoter regions.
```{r Checking for rejected percentage}
sum(!is.na(ALLpromotorCov))
rejectTotal <- (1-(sum(!is.na(ALLpromotorCov1))/sum(!is.na(ALLpromotorCov)))) 
reject <- (1-(sum(!is.na(ALLCov1Gen))/sum(!is.na(ALLCovGen))))
rejectTotal
reject
```


**Pairing coverage and Beta-values**

First a matrix containing only Beta-values is created under the name of ALLpromotorBeta.
```{r Creating matrix for Beta-values}
ALLpromotorBeta <- ALLpromotor[,c(5:14)]
```

Since a NA in the coverage means there is no data backing up the Beta-value the latter is also converted to NA. This is done in order to prevent incorporation of not significant Beta-values into downstream analysis.
```{r Pairing coverage and Beta-value NAs}
for(i in 1:nrow(ALLpromotorCov1)){
     for(j in 1:ncol(ALLpromotorCov1)){
         if(is.na(ALLpromotorCov1[i,j])){
             ALLpromotorBeta[i,j] <- NA
         }
     } 
}
```


**Elimination of all regions with more than 3 NA in Beta-value in both patient groups**

Any gene region with more than 3 NA for any patient group in the Beta-value coloumns is removed due to its low amount of information it can provide for testing of differentially methylated regions. Therefore if there is only one patient or none with a Beta-value for a specific gene region the latter is removed since the input data is not significant enough to be used in the downstream analysis.
```
for (i in 1:nrow(ALLpromotorBeta)) {
  for(j in  1:ncol(ALLpromotorBeta)){
    if(sum(is.na(ALLpromotorBeta[i,c(1:5)]))>=4){
      ALLpromotorBeta <- ALLpromotorBeta[-i,]
    }
  }
}
```

After execution of this command around 7.9 % of the promoter regions were removed due to 4 or 5 NA in healthy patients Beta-values.
```
for (i in 1:nrow(ALLpromotorBeta)) {
  for (j in 1:ncol(ALLpromotorBeta)) {
    if(sum(is.na(ALLpromotorBeta[i,c(6:10)]))>=4){
      ALLpromotorBeta <- ALLpromotorBeta[-i,]
    }
  }
}
```

After elimination through diseased patients another 1 % of promoter regions were removed. In total 8.9% (5225) of the promoter regions were removed. 


**Replacing residual NAs by the mean of the patient group**

Any remaining NAs are now replaced by the patient group's mean Beta-value in order to completely remove NAs in the Beta-values from our dataset.
```
for (i in 1:nrow(ALLpromotorBeta)) {       #replacement for healthy group
  for (j in 1:5) {
    if(is.na(ALLpromotorBeta[i,j])){
      ALLpromotorBeta[i,j] <- rowMeans(ALLpromotorBeta[i,c(1:5)], na.rm = TRUE) 
    }
  }
}
```

```
for (i in 1:nrow(ALLpromotorBeta)) {       #replacement for diseased group
  for (j in 6:10) {
    if(is.na(ALLpromotorBeta[i,j])){
      ALLpromotorBeta[i,j] <- rowMeans(ALLpromotorBeta[i,c(6:10)], na.rm = TRUE) 
    }
  }
}
```
The previous 4 code chunks were executed with a repeat loop until the number of NA in the dataset was equal to 0.
```
repeat{

  chunk1 #removal of gene regions with >=4 NA (healthy)
  
  chunk2 #removal of gene regions with >=4 NA (diseased)
  
  chunk3 #replacement of NA by mean of healthy patients
  
  chunk4 #replacement of NA by mean of diseased patients
  
  sum(is.na(ALLpromotorBeta))
  if(sum(is.na(ALLpromotorBeta)) == 0){
    break
  }
}
```

```{r Removing any NA, echo=FALSE}
repeat{
  for (i in 1:nrow(ALLpromotorBeta)) {
    for(j in  1:ncol(ALLpromotorBeta)){  # Alle Gene dim(start) = 59812
      if(sum(is.na(ALLpromotorBeta[i,c(1:5)]))>=4){
        ALLpromotorBeta <- ALLpromotorBeta[-i,]
      }
    }
  }

  for (i in 1:nrow(ALLpromotorBeta)) {
    for (j in 1:ncol(ALLpromotorBeta)) {       # Alle Gene dim(start) = 55111
      if(sum(is.na(ALLpromotorBeta[i,c(6:10)]))>=4){
        ALLpromotorBeta <- ALLpromotorBeta[-i,]
      }
    }
  }

  for (i in 1:nrow(ALLpromotorBeta)) {       # Healthy data
    for (j in 1:5) {
      if(is.na(ALLpromotorBeta[i,j])){
       ALLpromotorBeta[i,j] <- rowMeans(ALLpromotorBeta[i,c(1:5)], na.rm = TRUE) 
      }
    }
  }
  
  for (i in 1:nrow(ALLpromotorBeta)) {       # Diseased data
    for (j in 6:10) {
     if(is.na(ALLpromotorBeta[i,j])){
        ALLpromotorBeta[i,j] <- rowMeans(ALLpromotorBeta[i,c(6:10)], na.rm = TRUE) 
      }
    }
  }
  sum(is.na(ALLpromotorBeta))
  if(sum(is.na(ALLpromotorBeta)) == 0){
    break
  }
}
```


**Preparation for normalization**

In order to prevent mathematical errors during normalization for the values 0 and 1 they are replaced by 0.00001 and 0.99999 respectively. If this step is not done the resulting M-values would equate to positive and negative infinity.
```{r Preparing normalization}
ALLpromotorBeta1 <- ALLpromotorBeta
for (i in 1:nrow(ALLpromotorBeta)) {       
  for (j in 1:ncol(ALLpromotorBeta)) {
    if((ALLpromotorBeta[i,j])==0){
      ALLpromotorBeta[i,j] <- 0.00001 
    }
    if((ALLpromotorBeta[i,j])==1){
      ALLpromotorBeta[i,j] <- 0.99999
    }
  }
}
```


**Normalization**

In our last step of quality control our Beta-values were converted to M-values.
M-values provide a much better performance for detection rate and true positive rate at highly methylated and highly umethylated CpG sites. The Beta-values have a higher statistical spread (heteroscedasticity) outside the middle methylation range. So the M-Values provide a more accurate estimation of the methylation status (Pan Du *et al.*, 2010). The M-values have no direct biological interpretation, whereas the Beta-values relate roughly to the methylation percentage. For this reason we also looked at the Beta-Values in later on analysis.

```{r Converting Beta- to M-values}
ALLMvalue <- ALLpromotorBeta                 
for (i in 1:nrow(ALLMvalue)) {
  for (j in 1:ncol(ALLMvalue)) {
    ALLMvalue[i,j] <- log2(ALLMvalue[i,j]/(1-ALLMvalue[i,j]))
     }
}
```



### Step2 - Principal component analysis

Before testing single promoters for differential methylation we need to reduce the amount of data through principal component analysis (PCA). The dataset it therefore represented through a number of principal components (PC) with the first PC representing the highest variance of the data. Each PC is influenced by any promoter in the dataset. If the associated value is high the promoter is important for the principal component. If the value is low this particular promoter is not very important for the PC. 
Upon plotting the variance of each PC the relevant PCs can be filtered by cutting off at a noticeable elbow in the plot (as seen later). An elbow in the plot represents a cutoff point after which the following PCs/promoters don't add much information to the analyis. Therefore any following components/promoters can be removed without losing significant amounts of information about the data.
Next the selected PCs are plotted against each other in order to check for the best PC to pick for downstream analysis. Usually this is the first PC (PC1). As a last step the variance of each promoter is sorted and plotted for the selected PC. Here an elbow is visible again which marks the significant promoters on its left side (high variance) and the not significant ones to its right (low variance). Finally the significant genes are isolated and proceed to the next major step.


**Performing PCA and selection of relevant PCs**

First the actual PCA is executed.
```{r Performing PCA}
library(ggplot2)
pca <- prcomp(t(ALLMvalue))
```


Next the variance of the PCs are plotted to determine the elbow and therefore the relevant PCs. 
```{r Plotting variance of PCS}
pVar <- (pca$sdev)^2
plot(pVar, type = "l",main = "Plot of PCA variance", xlab = "PCs")
```

In our case the cutoff was at PC3 so we kept the first three principal components for the following steps.


**Plotting the PCs against each other**

In order to ensure the significance of the differential methylation we nedded to check for batch effects. We did this visually through plotting and statistically through several tests. Our samples did not show any significant batcheffect except for the tissue type since the control group contained tonsil and bone marrow samples. For this reason most analyses were performed without the tonsil samples. 


```{r Adding batch effect columns}
# ggPlot
  # Betrachtung der Patienten
  samples <- c(rep("h",5),rep("d",5)) # Erstellen einer weiteren Spalte im Datensatz für Farbigkeit
  tissue <- c(rep("tonsil",3),rep("bone marrow",7))
  provider <- c(rep("J.I.Martin-Subero",3),rep("J.Wiemels",2),rep("A.Bergmann",5))
  date <- c("Aug2013","Mar2015","Aug2015","Aug2013","Mar2015","May2016","May2016",rep("Jul2016",3))
```

```{r}
  pcax12 <- data.frame(pca$x[,c(1,2)]) # Matrix zu data.frame damit er damit arbeiten kann (ggplot)
  ggplot(pcax12, aes(x = PC1, y = PC2, colour = samples)) + geom_point() # Erzeugen ggPlot mit farbigen Punkten für PC12, PC13, PC23

  # Alternativ aber hier keine x-/y-Achsenbeschriftung
  # pcx12 <- data.frame(sample=rownames(p$x), X=p$x[,1], Y=p$x[,2]) 
  # ggplot(pcx12, aes(X, Y, group = sample)) + geom_point(aes(color = samples))
  
  # Problem mit Daten: Factor nicht numeric (manuelle Erstellung pcax12-dataframe)
  PC1 <- as.numeric(pca$x[,1])
  PC2 <- as.numeric(pca$x[,2])
  pcax12 <- cbind(PC1,PC2)
  pcax12 <- data.frame(pcax12)
  ggplot(pcax12, aes(x = PC1, y = PC2, colour = samples)) + geom_point()
  
  # Betrachtung der Batch-Effekt: Disease/Healthy, Tissue, Provider, Date
  batcheffect1 <- ggplot(pcax12, aes(x = PC1, y = PC2, colour = samples, shape = provider, size = tissue)) 
  batcheffect1 + geom_point()
  batcheffect2 <- ggplot(pcax12, aes(x = PC1, y = PC2, colour = date, shape = provider, size = tissue))
  batcheffect2 + geom_point()

# Betrachtung der Gene
pcar12 <- data.frame(pca$rotation[,c(1,2)])
ggplot(pcar12, aes(x = PC1, y = PC2)) + geom_point()
     
  # Filtern der Gene mit größter Varianz
  pcaRotVar <- apply(pca$rotation, 1, var) # Varianz der GenPCA
  pcaRotVarsort <- sort(pcaRotVar, decreasing = TRUE) # Sortierung der Varianz absteigen
  plot(pcaRotVarsort, type = "l", main = "Plot der Genvarianz", xlab = "Genes", ylim = c(0,0.0005), xlim = c(0,2000)) # Schauen wo Elbow (hier bei ca. 700)

  # Filtern der Gene nach Loading
  pcaRotAbs <- abs(pca$rotation[,1]) # Betrag der Loadings berechnen
  pcaRotAbssort <- sort(pcaRotAbs, decreasing = TRUE) # Loadings absteigend sortieren
  plot(pcaRotAbssort, type = "l", main = "Plot der PC1 Loading", xlab = "Genes", ylim = c(0,0.06), xlim = c(0,1000)) # Elbow bei unter 50 Genen (seltsam) 

  # Genposition nach Varianz ordnen (damit es auf die Methylierungsdaten übertragen werden kann)
  pcaRotVar <- order(pcaRotVar[,1], decreasing = TRUE) # Ordnen der Genpositionen nach Varianz (wie pcaRotVarsort)
  
  # Löschen der Gene hinter Varianz-elbow  
  pcaRotVar <- pcaRotVar[-c(701:54609)] # Alle Gene hinter Elbow (kleinere Varianz als 0.0001388)
  ALLMvalueRemain <- ALLMvalue[pcaRotVar,] # Extrahieren der Mvalues nach den verbliebenen Genpositionen
  ALLMvalueRemain <- rbind(ALLMvalueRemain,ALLMGen) # Miteinbinden der relevanten Gene (sind zuvor nicht im Dataframe gewesen)
      # ALLMvalueRemain: Enthält alle Gene/Patienten nach PCA
```

**Batcheffects**
```{r}
# Statistical Tests
pca_remain <- prcomp(t(ALLMvalueRemain)) 
  # Wilcoxon Rank Test
  wilcoxon_p1 <- wilcox.test(pca_remain$x[,1] ~ sample_annotation$TISSUE_TYPE) # Tissuetype von PC1-3 
  wilcoxon_p2 <- wilcox.test(pca_remain$x[,2] ~ sample_annotation$TISSUE_TYPE)
  wilcoxon_p3 <- wilcox.test(pca_remain$x[,3] ~ sample_annotation$TISSUE_TYPE)

  wilcoxon_p4 <- wilcox.test(pca_remain$x[,1] ~ sample_annotation$DISEASE) # Disease PC1-3 
  wilcoxon_p5 <- wilcox.test(pca_remain$x[,2] ~ sample_annotation$DISEASE)
  wilcoxon_p6 <- wilcox.test(pca_remain$x[,3] ~ sample_annotation$DISEASE)

  # Kruskal Wallis Test
  sample_annotation$BIOMATERIAL_PROVIDER <- as.factor(sample_annotation$BIOMATERIAL_PROVIDER) # Sonst Error Gruppenlevel muss endlich sein
  kruskal_p1 <- kruskal.test(pca_remain$x[,1] ~ sample_annotation$BIOMATERIAL_PROVIDER) # Provider PC1-3
  kruskal_p2 <- kruskal.test(pca_remain$x[,2] ~ sample_annotation$BIOMATERIAL_PROVIDER)
  kruskal_p3 <- kruskal.test(pca_remain$x[,3] ~ sample_annotation$BIOMATERIAL_PROVIDER)

  sample_annotation$FIRST_SUBMISSION_DATE <- as.factor(sample_annotation$FIRST_SUBMISSION_DATE)
  kruskal_p4 <- kruskal.test(pca_remain$x[,1] ~ sample_annotation$FIRST_SUBMISSION_DATE) # Date PC1-3 
  kruskal_p5 <- kruskal.test(pca_remain$x[,2] ~ sample_annotation$FIRST_SUBMISSION_DATE)
  kruskal_p6 <- kruskal.test(pca_remain$x[,3] ~ sample_annotation$FIRST_SUBMISSION_DATE)


# Visualisation
  # Heatmap
  Tissue_Type <- c(0.03,0.017,0.83)
  Disease <- c(0.008,0.69,1)
  Provider <- c(0.097,0.097,0.458)
  Submission_Date <- c(0.086,0.617,0.726)
  P_values <- rbind(Tissue_Type, Disease, Provider, Submission_Date) # Erstellen matrix mit p-values
  heatmap(P_values, main = "Heatmap P-values", Colv = NA, Rowv = NA, col = cm.colors(256)) # Heatmap der p-values
  # Levelplot
  library(lattice)
  levelplot(t(P_values), xlab = "PCs", ylab = "Batches", main = "Levelplot P-values") # Levelplot der p-values
  levelplot(t(ALLMvalueRemain[,c(1:3)]), xlab = "PCs", ylab = "Patients", main = "Levelplot PCA") # Levelplot der PCA$x-Werte
  pca_remain_x_abs <- abs(pca_remain$x[,c(1:3)])
  levelplot(t(pca_remain_x_abs), xlab = "PCs", ylab = "Patients", main = "Plot PCA Betrag") # Levelplot des Betrags der PCA$x-Werte
  
# Loading plotten 
pca_rot_PC1 <- pca_remain$rotation[,1] # Verwenden der PC1
pca_rot_PC1_Abs <- abs(pca_rot_PC1) # Loading Betrag berechnen
pca_rot_PC1_sort <- sort(pca_rot_PC1_Abs, decreasing = TRUE) # Loading sortieren
plot(pca_rot_PC1_sort, type = "l",main = "Loading plot PC1", ylab = "Loading", xlab = "Genes")

# K-Means (Silhouette-Plot)
M_km <- kmeans(t(x = ALLMvalueRemain), centers = 2, nstart = 10)
M_km
D <- dist(t(ALLMvalueRemain)) # Distanzmatrix
library(cluster)
silh <- silhouette(M_km$cluster,D) # Silhouettenplot der Patienten
plot(silh, ylab = "Patient") # 2 Cluster erkennbar
```



###Step3 - Testing for differentially methylated regions (DMR)

```{r}
# T-test
  Mvalues_noTO <- ALLMvalueRemain[,c(4:10)] # Tonsil-Patienten herausgenommen
  Mvalues_healthy <- Mvalues_noTO[,c(1,2)] # Matrix splitten in healthy und disease
  Mvalues_disease <- Mvalues_noTO[,c(3:7)]

t_test <- c(1:709) # Erstellung Liste für p-values

for (i in 1:length(t_test)) {
    t_test[i] <- t.test(Mvalues_healthy[i,],Mvalues_disease[i,])$p.value
}
  # Adjust p-value multiple comparison
  p_correction <- p.adjust(t_test, method = "holm", n = length(t_test))
  p_correction <- order(p_correction) # Korrigierte p-values absteigend ordnen -> Position wird ausgegeben
  t_test_order <- order(t_test) # Aufsteigend ordnen -> Position wird angegeben (Gleiche Anordnung wie bei p_correction)
  t_test_sort <- sort(t_test) # Aufsteigend ordnen -> Wert wird angegeben

t_test20 <- t_test_order[c(1:20)] # Erste 20 Werte behalten
ALLMvalueRemain20 <- ALLMvalueRemain[t_test20,] # Dataframe mit verbliebenene M-values (20)
ALLBetaRemain20 <- ALLpromotorBeta[t_test20,]   # Dataframe mit verbliebenen Beta-values (20)

threshold <- sum(t_test_sort <= 0.05) # Threshold auf p = 0.05 gesetzt
t_test_threshold <- t_test_order[c(1:threshold)]
ALLMvalueRemain_threshold <- ALLMvalueRemain[t_test_threshold,] # Dataframe mit verbliebenene M-values (99)
ALLBetaRemain_threshold <- ALLpromotorBeta[t_test_threshold,]   # Dataframe mit verbliebenen Beta-values (99)

# Visualisierung (Levelplot)
library(lattice)
  # Beta-value 20 verbliebene Genes
  ALLBetaRemain20_plot <- as.matrix(ALLBetaRemain20) # Konvertierung zur Matrix
  rownames(ALLBetaRemain20_plot) <- c(1:nrow(ALLBetaRemain20)) # Anpassen der Zeilennamen
  levelplot(ALLBetaRemain20_plot, xlab = "Genes", ylab = "Patients", main = "Levelplot Beta-values of 20 remaining Genes") 
  # M-value 20 verbliebene Genes
  for (i in 1:20) {
      for(j in 1:10){
          ALLMvalueRemain20[i,j] <- as.numeric(ALLMvalueRemain20[i,j]) # Nur für den Fall, dass Variablen nicht numerisch
      }
  }
  ALLMvalueRemain20_plot <- as.matrix(ALLMvalueRemain20) 
  rownames(ALLMvalueRemain20_plot) <- c(1:nrow(ALLMvalueRemain20))
  levelplot(ALLMvalueRemain20_plot, xlab = "Genes", ylab = "Patients", main = "Levelplot M-values of 20 remaining Genes") # Levelplot 20 verbliebenen M-values
  # Beta-values threshold p=0.05
  ALLBetaRemain_threshold_plot <- as.matrix(ALLBetaRemain_threshold)
  rownames(ALLBetaRemain_threshold_plot) <- c(1:nrow(ALLBetaRemain_threshold))
  levelplot(ALLBetaRemain_threshold_plot, xlab = "Genes", ylab = "Patients", main = "Levelplot Beta-values of remaining Genes") # Levelplot verbliebener beta-values hinter 0.05 threshold
  # M-values threshold p=0.05
  ALLMvalueRemain_threshold_plot <- as.matrix(ALLMvalueRemain_threshold)
  rownames(ALLMvalueRemain_threshold_plot) <- c(1:nrow(ALLMvalueRemain_threshold))
  levelplot(ALLMvalueRemain_threshold_plot, xlab = "Genes", ylab = "Patients", main = "Levelplot M-values of remaining Genes") # Levelplot verbliebener M-values hinter 0.05 threshold

```



###Step4 - Logistical regression

```{r}
# Logistical regression (threshold bei 99 Genes)
Lg_Mvalues <- data.frame(t(ALLMvalueRemain_threshold)) # Matrix transponieren damit Patienten auf y-Achse
Tumor <- factor(c(rep("0",5),rep("1",5)))
Lg_Mvalues <- cbind(Lg_Mvalues,Tumor) # Einbinden einer Spalte mit Bezeichnung ob Tumor ja/nein

LGpred <- as.matrix(c(1:10))  # Erstellung matrix für Ergebnisse
for (i in 1:nrow(ALLMvalueRemain_threshold)) {
     glm99 <- glm(Tumor ~ Lg_Mvalues[,i], family = "binomial", data
     = Lg_Mvalues) # Logistical regression einzelner Gene
     pred <- as.matrix(predict(glm99, type = "response")) # Anwendung zur Vorhersage
     LGpred <- cbind(LGpred,pred) # Matrix mit allen Ergebnissen der prediction
}
View(LGpred)
LGpred <- LGpred[,-1]

levelplot(t(LGpred), xlab = "Genes", ylab = "Patients", main = "Levelplot of logistical regression prediction") # Visualisierung
```

###Conclusion